---
title: "Ejemplo de k-medias"
css: estilos.css
date: '`r paste0("Última actualización: ", lubridate::now())`'
output: 
  html_document:
    toc: 1
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = T)
library(dplyr)
library(tidyr)
```

# Caracterización de segmentos.

Veamos otro ejemplo, utilizaremos la base de datos *Handwritten Digit Recognition Data*

This example is a character recognition task: classification of handwritten numerals. This problem captured the attention of the machine learning and neural network community for many years, and has remained a benchmark problem in the field.

Normalized handwritten digits, automatically scanned from envelopes by the U.S. Postal Service. The original scanned digits are binary and of different sizes and orientations; the images here have been deslanted and size normalized, resulting in 16 x 16 grayscale images (Le Cun et al., 1990).

The data are in two gzipped files, and each line consists of the digit id (0-9) followed by the 256 grayscale values.

There are 7291 training observations and 2007 test observations.



```{r}
library(ElemStatLearn)
library(parallel)
library(ggplot2)
```


```{r, cache=T}
dim(zip.train)
zip.train[1:6, 1:6]

dat <- zip.train[,-1]
nrow(dat)
ajustes.km <- mclapply(1:40, function(k){
  kmedias <- kmeans(dat, centers = k, nstart = 10, iter.max=40)    
  kmedias
}, mc.cores=6)
tot.within <- sapply(ajustes.km, function(aj){ aj$tot.withinss})
qplot(1:length(tot.within), tot.within, geom='line') + geom_point()
```

Ahora perfilamos los segmentos según los dígitos que contienen:

```{r}
grupos <- ajustes.km[[25]]$cluster
tabla <- table(ajustes.km[[25]]$cluster, zip.train[,1])
tabla
round(100*prop.table(tabla, margin=2))
round(100*prop.table(tabla, margin=1))

```


¿Cómo es el segmento 15?

```{r}
digits <- vector(length=10, mode="list")
names(digits) <- 0:9
rows <- sample(which(grupos==15),10)
for (j in 0:9) {
    digits[[j+1]] <- do.call("cbind", lapply(as.list(rows[[j+1]]), 
                       function(x) zip2image(zip.train, x)) )
}
im <- do.call("rbind", digits)
image(im, col=gray(256:0/256), zlim=c(0,1), xlab="", ylab="" )   

```


